Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-05-27 analysis/paper/paper.qmd
revise code and text

Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012]. MNI determines the actual numbers of individual animals recovered from remains, while NISP offers a broader overview by considering all fragmented bone assemblages. Despite criticism for potentially overestimating the number of animals, NISP can better capture variation in species abundance with less effect from sample size [@Grayson1984; @Lyman2008]. Bone fragments identified to the class level (e.g., Mammalia, Aves) were included in the overall NISP count to reflect the full scope of vertebrate remains. However, to enable more detailed taxonomic comparisons, indeterminate Mammalia specimens were excluded from further analysis to focus on bones identified to at least the family level. Deer antlers detached from the cranium were not considered in the quantification because they were less likely to be associated with dietary practices. To reduce biases resulting from differential number and recovery of skeletal elements among animal classes, NISP values were normalized by dividing the NISP by the average number of skeletal elements in a complete individual of that class [@Grayson2004].

To identify generalized patterns of animal consumption, measures of taxonomic abundance and ubiquity were applied. Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their use [@Mckechnie2016; @Vanderwarker2010]. In this study, ubiquity values are expressed as the percentage of units in which a given taxon is present relative to the total number of units per cultural layer. Given that mammals were the most commonly found animal resource, mammal bones identified to the family level were categorized individually for further discussion and compared in terms of their ubiquity with other animal classes.

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                     "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~
                                     "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
                                     "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~
                                     "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6,
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>%
  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100

buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
```

The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts). Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%). To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category. Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

>>>>

Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012]. MNI determines the actual numbers of individual animals recovered from remains, while NISP offers a broader overview by considering all fragmented bone assemblages. Despite criticism for potentially overestimating the number of animals, NISP can better capture variation in species abundance with less effect from sample size [@Grayson1984; @Lyman2008]. Bone fragments identified to the class level (e.g., Mammalia, Aves) were included in the overall NISP count to reflect the full scope of vertebrate remains. However, to enable more meaningful taxonomic comparisons, abundance analysis focused primarily on specimens identified to at least the family level, excluding indeterminate Mammalia. Deer antlers detached from the cranium were not considered in the quantification because they were less likely to be associated with dietary practices. To reduce biases resulting from differential number and recovery of skeletal elements among animal classes, NISP values were normalized by dividing the NISP by the average number of skeletal elements in a complete individual of that class [@Grayson2004].

The measure of ubiquity was applied to identify generalized patterns of animal consumption. Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their use [@Mckechnie2016; @Vanderwarker2010]. In this study, ubiquity values are expressed as the percentage of units in which a given taxon is present relative to the total number of units per cultural layer. Given that mammals were the most commonly found animal resource, mammal bones identified to the family level were categorized individually for further discussion and compared in terms of their ubiquity with other animal classes.

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "Large cervidae", "Cervidae large size")) %>%
  mutate(taxa = str_replace(taxa, "Testudines", "Geoemydidae")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cervus") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                     "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~
                                     "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
                                     "Ray-finned fish", str_detect(taxa, "Geo") ~ "Turtle",
                                   str_detect(taxa, "Cervidae") ~ "Deer", taxa == "Sus scrofa domestica" ~
                                     "Domestic pig", taxa == "Aves" ~ "Bird")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "munt") ~ 3, str_detect(taxa, "Sus") ~ 4, 
                           str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 5,  str_detect(taxa, "Rat") ~ 6,
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Geo") ~ 8,  
                           str_detect(taxa, "Ave") ~ 9,  str_detect(taxa, "large m") ~ 10, 
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(head(., 13), across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total determinate"))) %>%
  bind_rows(summarise(.[14:16, ], across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total indeterminate"))) %>%
  bind_rows(summarise(head(., 16), across(where(is.numeric), sum), 
                      across(where(is.character), ~"Total"))) %>%
  #bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
  mutate(`Weight (g)` = round(`Weight (g)`, 1)) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|str_detect(`Common name`, "Total"), "-", `Common name`)) %>%
  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
```

The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which were identified to at least the family level (@tbl-counts). Among the mammal bones identified to the family level, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%). To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category. Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
